Variable Selection in Convex Function Estimation

### 2013-10-04 tex/scam/icml14/optimization_icml.tex
minor edits

We now consider the following nonparametric regression problem
\begin{equation}\nonumber
          Y_{i} = f(\bds{x}_{i}) + \epsilon_{i} = 
                  \sum_{k=1}^{p}f_{k}(x_{ki}) + \epsilon_{i} \quad i=1,2,\cdots,n
\end{equation}
where $\bds{x}_{i}\in\mathbb{R}^{p}$ is the covariate, $Y_{i}$ is the
response and $\epsilon_{i}$ is mean zero noise. The regression function $f(\cdot)$ is the summation of 
functions $f_{k}(\cdot)$ in each variable dimension.  
We impose an additional constraint that each $f_{k}(\cdot)$ is 
a univariate convex function, which can be represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      h_{kj} \geq h_{ki} + \beta_{ki}(x_{kj}-x_{ki}) \quad (\forall i,j)
\end{equation}
where $h_{ki}\coloneqq f_{k}(x_{ki})$ and $\beta_{ki}$ is the
subgradient at point $x_{ki}$. We apparently need $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints, which is computationally
expensive for large scale problems.  In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to our optimization
program:
\begin{equation}
\begin{split}
       \min_{\bds{h},\bds{\beta},\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( Y_{i}-\sum_{k=1}^{p}h_{ki} - \mu \Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty} \\
       \textrm{subject to} &\;\; h_{k(i+1)} = h_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}), \\
                     & \;\; \sum_{i=1}^{n}h_{ki}=0,\\
                     & \;\; \beta_{k(i+1)} \geq \beta_{k(i)} \ (\forall k, i)
\end{split}
\label{np}
\end{equation}
Here $\{(1),(2),\ldots,(n)\}$ is a reordering of $\{1,2,\ldots,n\}$ such that $x_{k(1)}\leq{}x_{k(2)}\leq\cdots\leq{}x_{k(n)}$. We can solve for $\mu$ explicitly, as  
$\mu = \frac{1}{n} \sum_{i=1}^n Y_i = \bar{Y}$ which follows from the
KKT conditions
and the constraints $\sum_i h_{ki} = 0$.
It is easy to verify that the constraints in (\ref{np}) satisfy the supporting hyperplane constraints, as
\begin{align*}
  \forall j\geq{}i,\; h_{k(j)}-h_{k(i)}  &= 
            \sum\limits_{t=i}^{j-1}(h_{k(t+1)}-h_{k(t)}) \\
      & = \sum\limits_{t=i}^{j-1}\beta_{k(t)}(x_{k(t+1)}-x_{k(t)})\\ 
      & \geq \beta_{k(i)}\sum\limits_{t=i}^{j-1}(x_{k(t+1)}-x_{k(t)}) \\
      & = \beta_{k(i)}(x_{k(j)}-x_{k(i)}) \\
  \forall j<i,\;  h_{k(j)}-h_{k(i)} &=
                \sum\limits_{t=j}^{i-1}(h_{k(t)}-h_{k(t+1)}) \\
     &= \sum\limits_{t=j}^{i-1}\beta_{k(t)}(x_{k(t)}-x_{k(t+1)}) \\ 
     & \geq \beta_{k(i)}\sum\limits_{t=j}^{i-1}(x_{k(t)}-x_{k(t+1)})  \\
     &= \beta_{k(i)}(x_{k(j)}-x_{k(i)}).
\end{align*}
The $\ell_\infty/\ell_1$ penalty
$\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty}$ encourages group
sparsity of the vectors $\bds{\beta}_{k\cdot}$, and thus performs
variable selection.  We refer to this framework as the sparse convex
additive model (SCAM). While one can use
supporting hyperplanes to the epigraph as in \eqref{eq:outer}, 
SCAM uses the \emph{inner  piece-wise linear function}
that approximates the graph with secant lines. Notice that if we replace $\beta_{k(i+1)} \geq
\beta_{k(i)}$ with $\beta_{k(i+1)}=\beta_{k(i)}$, the optimization
reduces to the lasso.

@@

Where $\Delta_k$ is a $n\times n-1$ matrix such that $\Delta_k(i,j) = \Delta_{k(j)}(x_{k(i)})$ and $d_k = (d_{k(1)} ,..., d_{k(n-1)})$. We can now reformulate (\ref{np}) as an equivalent optimization program with only centering and positivity constraints:
\begin{align}
\min_{d_k, c_k}& \frac{1}{2n} 
       \Bigl\| Y - \bar{Y}\mathbf{1}_n - \sum_{k=1}^p 
             ( \Delta_k d_k - c_k \mathbf{1}_n) \Bigr\|_2^2 \nonumber \\
               & + \lambda_n \sum_{k=1}^p \|d_k\|_1   \label{opt:alternate_opt} \\
\trm{s.t. }  & d_{k(2)}, \ldots , d_{k(n-1)} \geq 0  	
               \qquad \trm{(convexity)} \nonumber \\ 
	& c_k = \frac{1}{n} \mathbf{1}_n^\tran \Delta_k d_k 	
               \qquad \trm{(centering)} \nonumber 
\end{align}
$\|d_k\|_1$ is not identical to $\|\bds{\beta}_{k\cdot}\|_{\infty}$, but it is easy to verify that $\|\bds{\beta}_{k\cdot}\|_{\infty} \leq \|d_k\|_1 \leq 2\|\bds{\beta}_{k\cdot}\|_{\infty}$.

>>>>

We now consider the following nonparametric regression problem
\begin{equation}\nonumber
          Y_{i} = f(\bds{x}_{i}) + \epsilon_{i} = 
                  \sum_{k=1}^{p}f_{k}(x_{ki}) + \epsilon_{i} \quad i=1,2,\cdots,n
\end{equation}
where $\bds{x}_{i}\in\mathbb{R}^{p}$ is the covariate, $Y_{i}$ is the
response and $\epsilon_{i}$ is mean zero noise. The regression function $f(\cdot)$ is the summation of 
functions $f_{k}(\cdot)$ in each variable dimension.  
We impose an additional constraint that each $f_{k}(\cdot)$ is 
a univariate convex function, which can be represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      h_{kj} \geq h_{ki} + \beta_{ki}(x_{kj}-x_{ki}) \quad (\forall i,j)
\end{equation}
where $h_{ki}\coloneqq f_{k}(x_{ki})$ and $\beta_{ki}$ is the
subgradient at point $x_{ki}$. We apparently need $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints, which is computationally
expensive for large scale problems.  In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to our optimization
program:
\begin{equation}
\begin{split}
       \min_{\bds{h},\bds{\beta},\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( Y_{i}-\sum_{k=1}^{p}h_{ki} - \mu \Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty} \\
       \textrm{subject to} &\;\; h_{k(i+1)} = h_{k(i)} + \beta_{k(i)}(x_{k(i+1)}-x_{k(i)}), \\
                     & \;\; \sum_{i=1}^{n}h_{ki}=0,\\
                     & \;\; \beta_{k(i+1)} \geq \beta_{k(i)} \ (\forall k, i)
\end{split}
\label{np}
\end{equation}
Here $\{(1),(2),\ldots,(n)\}$ is a reordering of $\{1,2,\ldots,n\}$ such that $x_{k(1)}\leq{}x_{k(2)}\leq\cdots\leq{}x_{k(n)}$. We can solve for $\mu$ explicitly, as  
$\mu = \frac{1}{n} \sum_{i=1}^n Y_i = \bar{Y}$ which follows from the
KKT conditions
and the constraints $\sum_i h_{ki} = 0$.
It is easy to verify that the constraints in (\ref{np}) satisfy the supporting hyperplane constraints, as
\begin{align*}
  \forall & j\geq{}i, \; h_{k(j)}-h_{k(i)}  = 
            \sum\limits_{t=i}^{j-1}(h_{k(t+1)}-h_{k(t)}) \\
      & = \sum\limits_{t=i}^{j-1}\beta_{k(t)}(x_{k(t+1)}-x_{k(t)})
       \geq \beta_{k(i)}\sum\limits_{t=i}^{j-1}(x_{k(t+1)}-x_{k(t)}) \\
      & = \beta_{k(i)}(x_{k(j)}-x_{k(i)}) \\
  \forall & j<i,\;  h_{k(j)}-h_{k(i)} =
                \sum\limits_{t=j}^{i-1}(h_{k(t)}-h_{k(t+1)}) \\
     & = \sum\limits_{t=j}^{i-1}\beta_{k(t)}(x_{k(t)}-x_{k(t+1)}) 
     \geq \beta_{k(i)}\sum\limits_{t=j}^{i-1}(x_{k(t)}-x_{k(t+1)})  \\
     &= \beta_{k(i)}(x_{k(j)}-x_{k(i)}).
\end{align*}
The $\ell_\infty/\ell_1$ penalty
$\sum_{k=1}^{p}\|\bds{\beta}_{k\cdot}\|_{\infty}$ encourages group
sparsity of the vectors $\bds{\beta}_{k\cdot}$, and thus performs
variable selection.  We refer to this framework as the sparse convex
additive model (SCAM). While one can use
supporting hyperplanes to the epigraph as in \eqref{eq:outer}, 
SCAM uses the \emph{inner  piece-wise linear function}
that approximates the graph with secant lines. Notice that if we replace $\beta_{k(i+1)} \geq
\beta_{k(i)}$ with $\beta_{k(i+1)}=\beta_{k(i)}$, the optimization
reduces to the lasso.

@@

Where $\Delta_k$ is a $n\times n-1$ matrix such that $\Delta_k(i,j) = \Delta_{k(j)}(x_{ki})$ and $d_k = (d_{k(1)} ,..., d_{k(n-1)})$. We can now reformulate (\ref{np}) as an equivalent optimization program with only centering and positivity constraints:
\begin{align}
\min_{d_k, c_k}& \frac{1}{2n} 
       \Bigl\| Y - \bar{Y}\mathbf{1}_n - \sum_{k=1}^p 
             ( \Delta_k d_k - c_k \mathbf{1}_n) \Bigr\|_2^2 \nonumber \\
               & + \lambda_n \sum_{k=1}^p \|d_k\|_1   \label{opt:alternate_opt} \\
\trm{s.t. }  & d_{k(2)}, \ldots , d_{k(n-1)} \geq 0  	
               \qquad \trm{(convexity)} \nonumber \\ 
	& c_k = \frac{1}{n} \mathbf{1}_n^\tran \Delta_k d_k 	
               \qquad \trm{(centering)} \nonumber 
\end{align}
$\|d_k\|_1$ is not identical to $\|\bds{\beta}_{k\cdot}\|_{\infty}$, but it is easy to verify that $\|\bds{\beta}_{k\cdot}\|_{\infty} \leq \|d_k\|_1 \leq 4\|\bds{\beta}_{k\cdot}\|_{\infty}$.
